Laurentia in the earliest Neoproterozoic: new paleomagnetic constraint from the Jacobsville Formation

### 2022-07-23 Manuscript/Manuscript.tex

\title{Laurentia's continued motion into the Neoproterozoic: new paleomagnetic constraint from the Jacobsville Formation}

\authors{Yiming Zhang\textsuperscript{1}, Blake Hodgin\textsuperscript{1}, Nicholas Swanson-Hysell\textsuperscript{1}, James Pierce\textsuperscript{1}, Anthony Fuentes\textsuperscript{1}}

\affiliation{1}{Department of Earth and Planetary Science, University of California, Berkeley, CA, USA}

@@

\begin{keypoints}
\item A new paleomagnetic pole from the ca. 992 Ma Jacobsville Formation 
\item Reversals put younger bounds on the normal Keweenawan superchron 
\item The new pole indicates the Grenville Loop could be younger than thought
\end{keypoints}

\begin{abstract}
The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and associated deposition of sediments during rift-related extension and post-rift thermal subsidence. Coeval with the cessation of rifting and a decrease of Laurentia's plate speed is the onset of the ca. 1090-980 Ma Grenvillian Orogeny along the eastern margin. The far-field stress of the orogeny is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation whose maximum depositional age has been constrained to be ca. 992 Ma. We develop a new paleomagnetic pole from red beds of the Jacobsville Formation along five stratigraphic sections in northern Michigan, USA. High-resolution thermal demagnetization experiments resolve detrital magnetic remanence that pass a fold test and two conglomerate tests. Our results confirm the occurrence of geomagnetic field reversals during the deposition of Jacobsville Formation, providing constraints for the maximum duration of the Keweenawan superchron. An inclination corrected paleomagnetic pole position from the Jacobsville Formation confirm continued plate motion of Laurentia across the equator in the late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the paleomagnetic poles developed from the Grenville Province most likely record coherent plate motion after major continental collision during the development of the supercontinent Rodinia.
% rather than recording a significant crustal shortening.
\end{abstract}

@@

\section*{Keywords}
Neoproterozoic\\
Laurentia  \\
Paleogeography\\
\

@@

The extensively studied paleomagnetic records of the ca. 1109 Ma to 1084 Ma North American Midcontinent Rift (i.e. the Keweenawan Track) provide crucial constraints on global paleogeography in the Mesoproterozoic \cite{Swanson-Hysell2019a}. As a result, the Keweenawan Track has become a central record for reconstructing the development of the ancient supercontinent Rodinia \cite{2021a}. As the rifting ceased and subsequently inverted following the onset of the Grenvillian Orogeny on Laurentia's eastern margin (Fig. \ref{fig:Geologic_map}), magmatism in the interior of Laurentia entered a period of quiescence that could have lasted for as long as $\sim$300 Myr until the emplacement of the ca. 780 Ma Gunbarrel large igneous province \cite{Harlan2003a, Mackinder2019a}. Due to this magmatic gap, Laurentia and its conjugate continents' paleogeographic configuration in the late Mesoproterozoic to mid-Neoproterozoic has remained largely uncertain \cite{Swanson-Hysell2021c}.

@@

When igneous rock records are lacking, paleomagnetic data recorded by sedimentary rocks (e.g. red beds) can provide useful insights into the ancient geomagnetic field and record paleogeographic positions of sedimentary basins. In the Midcontinent Rift, the paleomagnetic poles developed from the ca. 1075 Ma Nonesuch Formation and the ca. 1070 Ma upper Freda Formation of the Oronto Group sediments that overlie the Midcontinent Rift volcanics in southern Lake Superior area (Fig. \ref{fig:Geologic_map}) provide crucial constraints for Laurentia's latitude, orientation, and plate motion in the late Mesoproterozoic \cite{Henry1977a, Swanson-Hysell2019a, Swanson-Hysell2019b}.

On top of the Oronto Group sediments overlies the post-rift Jacobsville Formation which was partly deformed during the inversion of the Midcontinent Rift (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Hodgin2022a, DeGraff2022a}). The Jacobsville Formation also contains abundant red beds which offer the opportunity to further extend the reconstruction of the apparent polar wander path of Laurentia toward more recent times \cite{Hamblin1958a, Roy1978a}. U-Pb detrital zircon ages and U-Pb calcite ages from late-kinematic veins constrain the maximum depositional age of the Jacobsville Formation to be ca. 992 Ma \cite{Hodgin2022a}. Developing paleomagnetic data from the Jacobsville Formation will allow further development of Laurentia's paleogeographic evolution in the late Mesoproterozoic to early Neoproterozoic.

Challenges exist in interpreting sedimentary paleomagnetic records. Primary detrital remanence acquired during settling of magnetic particles in the water column can be masked as secondary magnetic minerals (e.g. pigmentary hematite) that precipitate and grow within pore spaces of host sedimentary rocks syn- to post-deposition. In red beds, it has been found that formation of the pigmentary hematite can significantly post-date the timing of sediment deposition, resulting in the chemical remanent magnetization being secondary, superimposed upon the primary remanent magnetization carried by detrital grains \cite{Butler1992a}. To isolate the remanence components of the red beds of the Jacobsville Formation, \citeA{Roy1974a, Roy1978a} adopted a method first applied by \citeA{Collinson1965c} to preferentially remove fine-grained pigmentary hematite through prolonged immersion in concentrated HCl acid whereas leaving the coarser detrital grains that are more resistant to dissolution in place. Studies applying paired acid etching demagnetization and thermal demagnetization show that the pigmentary hematite grains tend to have lower unblocking temperatures than detrital ones \cite{Tauxe1980a, Bilardello2010c}. High-resolution thermal demagnetization experiments with temperature intervals as small as 1-2 \textdegree C have shown to be effective in isolating detrital remanence from chemical remanence as coarser hematite grains tend to have higher unblocking temperatures \cite{Swanson-Hysell2019b}. In particular, a conglomerate test on a suite of fluvial intraclasts of the red beds in the Freda Formation shows that via thermal demagnetization one can isolate detrital magnetization vectors (which are reoriented in random directions in the clasts) from the secondary components carried by co-occurring authigenic hematite (which yield similar directions amongst clasts) \cite{Swanson-Hysell2019b}.

Paleomagnetism of red beds has also been challenged by the controversial issue of correcting for inclination shallowing \cite{King1955a, Tauxe2004b, Bilardello2016b}, which is necessary due to the rotation of detrital magnetic grains to be parallel to bedding caused by authigenic processes such as compaction during lithification. As a result, inclinations recorded by sediments often have shallower angles than the local field inclinations during the time of deposition. To estimate the amount of inclination shallowing in a given package of sedimentary rocks, \citeA{Tauxe2004b} developed a statistical model (i.e. the elongation/inclination method) that evaluates the deviation of the distribution of a large number ($>$100) of observed sedimentary paleomagnetic directional data with the predicted distribution given by a paleosecular variation model developed for the past 5 Myr (TK03) under the the geocentric axial dipole (GAD) hypothesis. Although the TK03 paleosecular variation model is based on data of relatively recent geomagnetic field, it has been shown to be compatible with data from the ca. 1.1 Ga lava flows of the Midcontinent Rift \cite{Tauxe2009a}. Further, \citeA{Pierce2022a} showed that the E/I method passes a field test for the 1.1 Ga Cut Face Creek Sandstone as the method successfully corrects the shallow paleomagnetic inclination of the sediments to that of the lava flows that bracket them. That study also further extended the method to better represent uncertainties associated with the amount of inclination shallowing when reporting sedimentary paleomagnetic pole positions by using the spherical elliptical Kent distribution \cite{Kent1982a} instead of the circular Fisher distribution \cite{Fisher1953a}.

In this study, we investigate the hematite-bearing, fine-grained sandstone to siltstone lithofacies along five stratigraphic sections of the Jacobsville Formation. High-resolution thermal demagnetization data on siltstone intraclasts that were eroded by fluvial processes and redeposited at two locations constrain the timing of hematite acquisition by revealing a primary component that formed prior to the erosion of the clasts within the depositional environment and a secondary component that formed following their redeposition. We then use a paleomagnetic fold test to further constrain the timing of detrital remanence acquisition in association with the structural inversion of the Midcontinent Rift. Finally, we develop a new inclination-corrected paleomagnetic pole for the Jacobsville Formation based on a large number of samples and investigate the paleo-continental configuration of Laurentia and its association with the development of the supercontinent Rodinia.

% The end of the ca. 1109 Ma to 1084 Ma intracontinental Midcontinent Rift magmatism and extension was marked by a transition to deposition of sedimentary rocks of the Oronto Group \cite{Ojakangas2001a} as the rift basin thermally subsided prior to the rift undergoing contractional deformation in the later stages of the Grenvillian orogeny \cite{Cannon1992b, Hodgin2022a}.

@@

% A leading tectonic model for the cessation of the Midcontinent Rift and extension is that it occurred due to far-field compressional stress associated with the ca. 1090-980 Ma Grenvillian orogeny \cite{Cannon1992a}, when the collision of Laurentia with its conjugate margins led to the development of a back-bulge basin, accommodating the deposition of the Jacobsville Formation sediments as the inversion of the Midcontinent Rift progressed. Recent re-examination of the maximum deposition age of the Jacobsville Formation using zircon U-Pb whole grain chemical abrasion isotope dilution thermal ionization mass spectrometry has updated the maximum deposition age of the Jacobsville Formation to be ca. 990 Ma \cite{Hodgin2022a}. Such age constraint is consistent with backbulge subsidence model \cite{DeCelles, 2012}, indicating that the Jacobsville Formation may have been deposited in a Grenvillian foreland basin system that resulted from lithospheric flexure induced by orogenic loading \cite{Rivers2012a}. The relatively young age of the Jacobsville Formation compared to other Midcontinent Rift rocks makes the Jacobsville Formation an crucial target for further reconstructing the global paleogeography in the late Proterozoic. For further refining the paleogeographic connection between the poles of the Grenville Loop and those of the Keweenawan Track.

@@

% The extensive paleomagnetic studies of the well-preserved late Mesoproterozoic to Neoproterozoic rocks of North America have provided the central record for global paleogeography reconstruction during the assembly and rearrangement of the postulated supercontinent Rodinia. Paleomagnetic studies of rocks from Laurentia have led to a series of paleomagnetic poles that form an apparent polar wander path (APWP) known as the ``Logan Loop" for the older high latitude poles at its apex that continues into the ``Keweenawan Track" of younger lower latitude poles that form a progression as the APWP heads toward the ``Grenville Loop" \citep{Swanson-Hysell2019a}. In particular, recent advances in pairing high-precision zircon U-Pb geochronology with high-quality paleomagnetic poles significantly improved the resolution of the Keweenawan Track and revealed Laurentia's rapid motion ($>$20 cm/yr) during the Midcontinent Rift magmatic activity from ca. 1110 Ma to ca. 1085 Ma \citep{Swanson-Hysell2019a}.

@@

% Paleomagnetic poles from the interior of Laurentia during the quiescence magmatism period is thus crucial for reconcile the discordant paleomagnetic poles of the Keweenawan Track and the Grenville Loop. The
% %Q: what are you using the poles from rocks that we know are from the Laurentia interior? (outside of the Grenville package?)

% This new age constraint on the Jacobsville provides crucial insight into the tectonic evolution of the Midcontinent Rift and the inversion of the extensional region into a compressional region due to the Grenville orogen. It also makes the Jacobsville sandstone a unique opportunity to investigate Laurentia's paleogeographical position at ca. 1 Ga as a long period of magmatic quiescence for $\sim$50 myr precedes Jacobsville deposition left a gap in Laurentia apparent polar wander path between the Keweenawan Track and the Grenville Loop (Fig. xxx).

% Jacobsville sedimentation was preceded by a long period of volcanic and tectonic quiescence and cratonic stability so that bedrock surfaces became blanketed by paleosol and a surface of chemically resistant debris was dominated by quartz and iron-formation.

% Previous paleomagnetic study on the red beds within the Jacobsville by \cite{Roy1978a} observed different components using AF, thermal, and chemical demagnetizations. Recognized dual polarities and recognized the complex remanence acquisition of red beds which lead to three isolated remanence components from the Jacobsville. However, that study did not recognize the problem of inclination shallowing and did not apply correction factor on the results. Moreover, the course demagnetization steps during thermal demagnetization would have had limited resolution for distinguishing the depositional remanent magnetization (DRM) and secondary chemical remanent magnetization (CRM) which have recently been shown to unblock over a narrow temperature window \cite{Swanson-Hysell2019b}.

@@

\subsection*{Structure}
The ca. 1109 Ma to 1084 Ma North American Midcontinent Rift is a major intracontinental rift system that extends over 2000 km through the Laurentia craton (Fig. \ref{fig:Geologic_map}). However, the rifting failed to separate Laurentia into two continents. As its magmatism waned and thermal subsidence progressed in the rift basin, the $>$4 km of ca. 1085-1050 Ma Oronto Group sediments \cite{Fairchild2017a, Swanson-Hysell2019a} were deposited conformably on top of the rift volcanics (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1992a}). Subsequently, the rift was inverted as the ca. 1090-980 Ma Grenvillian Orogeny commenced on the eastern margin of Laurentia (Fig. \ref{fig:Geologic_map}), eventually resulting in the propagation of far-field compressional stress deep into the interior of Laurentia \cite{Cannon1993a, Cannon1994a}.

During the inversion, Paleoproterozoic and Archean crust was uplifted via thrust faults such as the Marenisco fault and the Pelton Creek fault, forming the crustal-scale Montreal River monocline (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1993a}). The southern part of the post-rift Jacobsville Formation overlies an erosional angular unconformity that developed on lithologies that were exhumed through this earlier reverse motion (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Cannon1993a, Kalliokoski1982a}). More recent mapping of the Jacobsville Formation has also found that it can sometimes onlap the Midcontinent Rift volcanics within or adjacent to the Keweenaw Fault system around Bete Grise Bay on the Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}; \citeA{Tyrrell2019a, Mueller2021a}). Field observations find erosional contact and development of paleosol between the Jacobsville Formation and underlying volcanic rocks \cite{Hamblin1958a, Kalliokoski1975a}, and interpretations of seismic reflection data show that sedimentary rocks interpreted to be Jacobsville Sandstone lie unconformably on the Oronto Group \cite{Cannon1989a}. These observations are consistent with the interpretation that there was a prolonged period of sedimentation quiescence and net erosion between the end of Oronto Group deposition and onset of Jacobsville Formation deposition. Based on dates from Cannon et al 1993 (and deformation of the Freda Fm within the Montreal River monocline) paired with new dates on the Jacobsville Fm (Hodgin et al., 2022), this time period of quiescence may have lasted up to ~60 Myrs from ca. 1050 until ca. 990 Ma. The overall extent of the Jacobsville Formation has been mapped along the lake shore of the Keweenaw Peninsula from northern Michigan, USA, to the north side of Michipicoten Island and perhaps to northern Sault Ste Marie, Canada (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a}).

As far-field stresses from the Grenvillian Orogeny propagated into the interior of Laurentia, the previously thinned crust of the Midcontinent Rift was structurally inverted by crustal-scale thrusts, such as the Keweenaw Fault which extends $\sim$250km from ...refs here. Final inversion of the Keweenaw Fault during the Grenvillian Orogeny thrust Midcontinent Rift volcanics over the Jacobsville strata, which were folded in the footwall of the fault.  along much of the Keweenaw fault which extends $\sim$250 km from the tip of the Keweenaw Peninsula in northern Michigan southeastward to a termination in northeastern Wisconsin (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1996a, Cannon2001a, DeGraff2022a}). Syndepositional faulting that took place during Jacobsville deposition is inferred from ... rift volcanic clasts where eroded into the Jacobsville basin at the same time when sediments from the older basement rocks along with distal input from the Grenville Province were deposited within the Jacobsville back-bulge basin system that in general wedges toward the northwest (Fig. \ref{fig:Geologic_map}; \cite{Hedgman2014a, Hodgin2022a}. Next to the footwall of the Keweenaw fault, folding has also been recorded by strata of the upper Jacobsville Formation in contact with the volcanics \cite{Cannon1993a}. One famous remnant of that late-stage deformation is the Natural Wall ravine where sub-vertical sandstone beds are preserved (Fig. \ref{fig:Geologic_map}; \cite{Hamblin1958a}). On the shore of Bete Grise Bay near the northeastern end of Keweenaw Peninsula, the Keweenaw fault is interpreted to have accommodated Grenville compressional stresses by forming a series of segmented en echelon thrusts \cite{Tyrrell2019a}. The timing of such faulting could be coeval with the deposition of the upper Jacobsville Formation, leading to the nonconformable onlapping of the Jacobsville sediments over the volcanics in addition to the thrust faulting relationship between the two units. Ca. 985 Ma late-kinematic calcite that precipitated within the hanging wall of the Keweenaw Fault and ca. 992 Ma maximum deposition age from detrital zircons of the Jacobsville Formation in the footwall constrain the timing of the final inversion of the Keweenaw Fault to be during the Rigolet phase of the Grenvillian Orogeny \cite{Hodgin2022a}.

\subsection{Sedimentology and general stratigraphy}

% It has been interpreted that the Jacobsville Sandstone is equivalent to the Bayfield Group in northern Wisconsin, and the Fond Du Lac and Hinckley formations in Minnesota. Their shared lithofacies of basal conglomerate, cross-bedded fluvial fine to medium grained sandstone, and interbedded red micaceous siltstone could support the interpretation that these sedimentary rocks formed during a common period of post-rift basin formation following rift-related magmatism of the Midcontinent Rift. 
\begin{figure*}[h!]
\centering
\includegraphics[width=0.9\textwidth]{Field_photo.png}
\caption{\scriptsize Field photos of red beds of the Jacobsville Formation sampled in this study. At Baby Snake Creek and the Natural Wall ravine in northern Keweenaw Peninsula, intervals of red fine-grained sandstone beds can be found through the steeply dipping to overturned basal strata near the Keweenaw Fault (A,C) and also the nearly horizontal top strata (B,D). At Hungarian Falls, relatively thick intervals of fine-grained red sandstone to siltstone beds separated by hematite-depleted arenite to arkosic sandstones or conglomerates are found along a $\sim$100-meter-tall waterfall (E). Abundant reoriented siltstone rip-up clasts often exist within the conglomeratic layers above siltstone beds (F). Drill cores or block samples of these rip-up clasts are sampled for the paleomagnetic conglomerate test presented in Fig. \ref{fig:Intraclast_pmag}. At Agate Falls, nearly flat-lying red fissile shale beds are exposed due to waterfall incision (G). Subrounded detrital mica grains deposited parallel to the bedding plane are often associated with the siltstone facies. Another set of siltstone rip-up clasts were sampled from a conglomeratic layer near the basal exposure for the paleomagnetic conglomerate test. }
\label{fig:Field_photo}
\end{figure*}

>>>>

\title{Laurentia in the earliest Neoproterozoic: new paleomagnetic constraint from the Jacobsville Formation}

\authors{Yiming Zhang\textsuperscript{1}, Blake Hodgin\textsuperscript{1}, Nicholas Swanson-Hysell\textsuperscript{1}, James Pierce\textsuperscript{1,2}, Anthony Fuentes\textsuperscript{1}}

\affiliation{1}{Department of Earth and Planetary Science, University of California, Berkeley, CA, USA}
\affiliation{2}{Department of Earth and Planetary Sciences, Yale University, CT, USA}

@@

\begin{keypoints}
\item A new paleomagnetic pole from the ca. 990 Ma Jacobsville Formation \item Laurentia crossed the equator in the latest Mesoproterozoic to earliest Neoproterozoic
\item The new pole indicates the Grenville Loop could be much younger than thought
\end{keypoints}

\begin{abstract}
Coeval with the cessation of the ca. 1.1 Ga intracontinental Midcontinent Rift was the onset of the ca. 1090-980 Ma Grenvillian Orogeny along Laurentia's eastern margin. Far-field compressional stresses of the orogeny propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation whose maximum depositional age has been constrained to be ca. 990 Ma. We develop a new paleomagnetic pole from red beds of the Jacobsville Formation in northern Michigan, USA. High-resolution thermal demagnetization experiments resolve detrital remanent magnetizations that pass a conglomerate test. An inclination corrected paleomagnetic mean pole position from the Jacobsville Formation confirm continued plate motion of Laurentia across the equator in the late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the paleomagnetic poles developed from the Grenville Province most likely record coherent plate motion after major continental collision during the initial assembly of the supercontinent Rodinia.
\end{abstract}

@@

\section*{Keywords}
Rodinia\\
Laurentia  \\
Paleogeography\\
Neoproterozoic\\
\

@@

The extensively studied paleomagnetic records of the ca. 1109 Ma to 1084 Ma North American Midcontinent Rift (i.e. the Keweenawan Track) provide crucial constraints on the paleogeography of Laurentia in the Mesoproterozoic \cite{Swanson-Hysell2019a}. As a result, the Keweenawan Track has become a central record for reconstructing the initial development of the ancient supercontinent Rodinia \cite{2021a}. As the rifting ceased and subsequently inverted following the onset of the Grenvillian Orogeny on Laurentia's eastern margin (Fig. \ref{fig:Geologic_map}), magmatism in the interior of Laurentia entered a period of quiescence that could have lasted for as long as $\sim$300 Myr until the emplacement of the ca. 780 Ma Gunbarrel large igneous province \cite{Harlan2003a, Mackinder2019a}. Due to this magmatic gap, Laurentia and its conjugate continents' paleogeographic configuration in the late Mesoproterozoic to mid-Neoproterozoic has remained largely uncertain \cite{Swanson-Hysell2021c}.

@@

When igneous rock records are lacking, paleomagnetic data recorded by sedimentary rocks (e.g. red beds) can provide useful insights into the ancient geomagnetic field and record paleogeographic positions of sedimentary basins. In the Midcontinent Rift, the paleomagnetic poles developed from the ca. 1075 Ma Nonesuch Formation and the ca. 1070 Ma upper Freda Formation of the Oronto Group sediments that overlie the Midcontinent Rift volcanics in southern Lake Superior area (Fig. \ref{fig:Geologic_map}) provided crucial constraints for Laurentia's paleolatitude, orientation, and plate speed in the late Mesoproterozoic \cite{Henry1977a, Swanson-Hysell2019a, Rose2022a}.

On top of the Oronto Group sediments overlies the post-rift Jacobsville Formation which was partly deformed during the inversion of the Midcontinent Rift (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Hodgin2022a, DeGraff2022a}). U-Pb detrital zircon ages and U-Pb calcite ages from the Keweenaw fault where the Jacobsville Formation is deformed against thrusted rift volcanics constrain the maximum depositional age of the Jacobsville Formation to be ca. 990 Ma \cite{Hodgin2022a}. The Jacobsville Formation contains abundant hematite-rich red beds which bear potential to further extend the apparent polar wander path (APWP) of Laurentia toward more recent times \cite{Hamblin1958a, Roy1978a}. Developing paleomagnetic data from the Jacobsville Formation will allow further development of Rodinia's paleogeographic evolution in the late Mesoproterozoic to early Neoproterozoic.

Challenges exist in interpreting sedimentary paleomagnetic records. Primary detrital remanence acquired through settling and deposition of magnetic particles in the water column can be masked by secondary remanence acquired through precipitation and growth of new minerals within pore spaces of host sedimentary rocks syn- to post-deposition (e.g. pigmentary hematite). In red beds, it has been found that formation of the pigmentary hematite can significantly post-date the timing of sediment deposition, resulting in a chemical remanent magnetization (CRM) superimposed upon primary detrital remanent magnetization (DRM) carried by detrital grains \cite{Butler1992a}. To isolate the remanence components of the red beds of the Jacobsville Formation, \citeA{Roy1974a, Roy1978a} adopted a method first applied by \citeA{Collinson1965c} to preferentially remove fine-grained pigmentary hematite through prolonged immersion in concentrated HCl acid whereas leaving the coarser detrital grains that are more resistant to dissolution in place. Studies applying paired acid etching demagnetization and thermal demagnetization show that the pigmentary hematite grains tend to have lower unblocking temperatures than detrital ones \cite{Tauxe1980a, Bilardello2010c}. High-resolution thermal demagnetization experiments with temperature intervals as small as 1-2 \textdegree C have shown to be effective in isolating detrital remanence from chemical remanence as coarser hematite grains tend to have higher unblocking temperatures closer to the Ne\'el temperature of hematite \cite{Swanson-Hysell2019b}. In particular, a conglomerate test on a suite of fluvial intraclasts of the red beds in the Freda Formation shows that via thermal demagnetization one can isolate detrital magnetization vectors (which are reoriented in random directions in the clasts) from the secondary components carried by co-occurring authigenic hematite (which yield similar directions amongst clasts) \cite{Swanson-Hysell2019b}.

Paleomagnetism of red beds has also been challenged by the controversial issue of correcting for inclination shallowing \cite{King1955a, Tauxe2004b, Bilardello2016b}. Due to rotation of detrital magnetic grains caused by authigenic processes such as compaction during lithification, inclinations recorded by sediments often have shallower angles than the local field in which they are deposited. To estimate the amount of inclination shallowing in a given package of sedimentary rocks, \citeA{Tauxe2004b} developed a statistical model (i.e. the elongation/inclination [E/I]method) that evaluates the deviation of the distribution of a large number ($>$100) of observed sedimentary paleomagnetic directional data with the predicted distribution given by a paleosecular variation model developed for the past 5 Myr (TK03) under the the geocentric axial dipole (GAD) hypothesis. Although the TK03 paleosecular variation model is based on data of relatively recent geomagnetic field variations, it has been shown to be compatible with data from the ca. 1.1 Ga lava flows of the Midcontinent Rift \cite{Tauxe2009a}. Further, \citeA{Pierce2022a} showed that the E/I method passes a field test for the 1.1 Ga Cut Face Creek Sandstone in the Midcontinent Rift as it successfully corrects the shallow paleomagnetic inclinations of the sediments to that of the North Shore Volcanic Group lava flows that bracket them. That study also further extended the method to better represent uncertainties associated with the amount of inclination shallowing when reporting sedimentary paleomagnetic pole positions by using the spherical elliptical Kent distribution \cite{Kent1982a} instead of the circular Fisher distribution \cite{Fisher1953a}.

In this study, we investigate the hematite-bearing, fine-grained sandstone to siltstone lithofacies along five stratigraphic sections of the Jacobsville Formation on Keweenaw Peninsula, Michigan, USA. High-resolution thermal demagnetization isolate primary DRM from secondary CRM with the DRM passing a conglomerate test. We then use a paleomagnetic fold test to further constrain the timing of DRM acquisition in association with the structural inversion of the Midcontinent Rift. Finally, we develop a new inclination-corrected paleomagnetic pole for the Jacobsville Formation based on a large number of samples and investigate the paleo-continental configuration of Laurentia in the late Mesoproterozoic to early Neoproterozoic.

% The end of the ca. 1109 Ma to 1084 Ma intracontinental Midcontinent Rift magmatism and extension was marked by a transition to deposition of sedimentary rocks of the Oronto Group \cite{Ojakangas2001a} as the rift basin thermally subsided prior to the rift undergoing contractional deformation in the later stages of the Grenvillian orogeny \cite{Cannon1990b, Hodgin2022a}.

@@

% A leading tectonic model for the cessation of the Midcontinent Rift and extension is that it occurred due to far-field compressional stress associated with the ca. 1090-980 Ma Grenvillian orogeny \cite{Cannon1990a}, when the collision of Laurentia with its conjugate margins led to the development of a back-bulge basin, accommodating the deposition of the Jacobsville Formation sediments as the inversion of the Midcontinent Rift progressed. Recent re-examination of the maximum deposition age of the Jacobsville Formation using zircon U-Pb whole grain chemical abrasion isotope dilution thermal ionization mass spectrometry has updated the maximum deposition age of the Jacobsville Formation to be ca. 990 Ma \cite{Hodgin2022a}. Such age constraint is consistent with backbulge subsidence model \cite{DeCelles, 2012}, indicating that the Jacobsville Formation may have been deposited in a Grenvillian foreland basin system that resulted from lithospheric flexure induced by orogenic loading \cite{Rivers2012a}. The relatively young age of the Jacobsville Formation compared to other Midcontinent Rift rocks makes the Jacobsville Formation an crucial target for further reconstructing the global paleogeography in the late Proterozoic. For further refining the paleogeographic connection between the poles of the Grenville Loop and those of the Keweenawan Track.

@@

@@

@@

\subsection*{Structure}
The ca. 1109 Ma to 1084 Ma North American Midcontinent Rift is a major intracontinental rift system that extends over 2000 km through the Laurentia craton (Fig. \ref{fig:Geologic_map}). However, the rifting failed to separate Laurentia into two continents. As its magmatism waned and thermal subsidence progressed in the rift basin, the $>$4 km of ca. 1085-1050 Ma Oronto Group sediments \cite{Fairchild2017a, Swanson-Hysell2019a} were deposited conformably on top of the rift volcanics (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1989a}). Subsequently, the rift was inverted as the ca. 1090-980 Ma Grenvillian Orogeny commenced on the eastern margin of Laurentia (Fig. \ref{fig:Geologic_map}), eventually resulting in the propagation of far-field compressional stresses deep into the interior of Laurentia \cite{Cannon1993a, Cannon1994a}.

During the rift inversion, Paleoproterozoic and Archean crust was uplifted via thrust faults such as the Marenisco fault and the Pelton Creek fault in the south, forming the crustal-scale Montreal River monocline (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1993a}). The southern part of the post-rift Jacobsville Formation overlies an erosional angular unconformity that developed on lithologies that were exhumed through this earlier reverse motion (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Cannon1993a, Kalliokoski1982a}). North to these faults, Midcontinent Rift volcanic rocks were thrusted and folded along the Keweenaw fault which extends $\sim$250km from northern Wisconsin to northern Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}; \citeA{DeGraff2022a}). Final inversion of the Keweenaw fault during the Rigolet phase of the Grenvillian Orogeny deformed part of the Jacobsville strata near the fault \cite{Hodgin2022a}. Recent mapping of the Jacobsville Formation also found that it can sometimes onlap the Midcontinent Rift volcanics within or adjacent to the Keweenaw Fault system near the northern end of the Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}; \citeA{Tyrrell2019a, Mueller2021a}).

Field observations find erosional contact and development of paleosol between the Jacobsville Formation and underlying volcanic rocks \cite{Hamblin1958a, Kalliokoski1975a}, and interpretations of seismic reflection data show that sedimentary rocks interpreted to be Jacobsville Formation lie unconformably on the Oronto Group \cite{Cannon1989a}. These data indicate that there was a prolonged period of sedimentation quiescence and net erosion between the end of Oronto Group deposition and onset of Jacobsville Formation deposition. Rb-Sr radiometric thermochronology data of \citeA{Cannon1993a} show the Montreal River monocline in which the Freda Formation and older basement rocks are deformed had been uplifted to mid to shallow crustal levels by ca. 1060 Ma. U-Pb dates of late kinematic calcite crystals that precipitated at ($\sim$50\textdegree C) in the hanging wall of the Keweenaw fault \cite{Hodgin2022a} indicate that by ca. 985 Ma the dominant inversion of Midcontinent Rift rocks along the Keweenaw fault has completed. Together, these data are consistent with the interpretation that the dominant inversion of the Midcontinent Rift basin happened during the ca. 1090-1020 Ma Ottawan phase of the Grenvillian Orogeny. They also indicate that the period of sedimentary quiescence in the rift basin may have lasted up to $\sim$70 Myrs from ca. 1060 to ca. 990 Ma.

Syndepositional faulting that took place during Jacobsville deposition is inferred based on the observation of rift volcanic clasts within the Jacobsville basin. In the footwall of the Keweenaw fault near central Keweenaw Peninsula, strata of the Jacobsville Formation in contact with the Midcontinent volcanics are often steeply folded with the conglomeratic strata containing volcanic clasts \cite{Cannon1993a}. On the shore of Bete Grise Bay near the northeastern end of Keweenaw Peninsula, the Keweenaw fault is interpreted to have accommodated Grenville compressional stresses by forming a series of segmented en echelon thrusts \cite{Tyrrell2019a}. The timing of such faulting could be coeval with the deposition of the upper Jacobsville Formation, leading to the nonconformable onlapping of the Jacobsville sediments over the volcanics in addition to the thrust faulting relationship between the two units.

% at the same time when sediments from the older basement rocks along with distal input from the Grenville Province were deposited within the Jacobsville back-bulge basin system that in general wedges toward the northwest (Fig. \ref{fig:Geologic_map}; \cite{Hedgman2014a, Hodgin2022a}.

\subsection{Sedimentary facies and general stratigraphy with interpretation}

% It has been interpreted that the Jacobsville Sandstone is equivalent to the Bayfield Group in northern Wisconsin, and the Fond Du Lac and Hinckley formations in Minnesota. Their shared lithofacies of basal conglomerate, cross-bedded fluvial fine to medium grained sandstone, and interbedded red micaceous siltstone could support the interpretation that these sedimentary rocks formed during a common period of post-rift basin formation following rift-related magmatism of the Midcontinent Rift. 
\begin{figure*}[h!]
\centering
\includegraphics[width=0.9\textwidth]{Field_photo.png}
\caption{\scriptsize Field photos of red beds of the Jacobsville Formation. At Baby Snake Creek and the Natural Wall ravine in northern Keweenaw Peninsula, intervals of red fine-grained sandstone to siltstone beds can be found through the steeply dipping to overturned basal strata near the Keweenaw fault (A,C) and also the nearly horizontal top strata (B,D). At Hungarian Falls, relatively thick intervals of red beds separated by hematite-depleted arenite to arkosic sandstones or conglomerates are found along a $>$100-meter-tall waterfall (E). Abundant reoriented siltstone rip-up clasts often exist within the conglomeratic layers above siltstone beds (F). Drill cores and block samples of these rip-up clasts were sampled for the paleomagnetic conglomerate test presented in Fig. \ref{fig:intraclast_pmag}. At Agate Falls, nearly flat-lying red fissile siltstone beds are exposed. (G). Subrounded detrital mica grains deposited parallel to the bedding plane are often associated with the siltstone facies. Another set of siltstone rip-up clasts were sampled from a conglomeratic layer near the basal exposure for the paleomagnetic conglomerate test. }
\label{fig:Field_photo}
\end{figure*}

The Jacobsville Formation is a fluvial sequence of feldspathic and quartzose conglomerates, sandstones, siltstones, and shales completely devoid of lava flows or cross-cutting igneous dikes although numerous  cross-cutting clastic dikes are present \cite{Hamblin1958a}. Sediments of the Jacobsville Formation is interpreted to be dominantly derived from the Oronto Group sediments and older basement rocks that were uplifted and deformed in the Montreal River monocline. The Jacobsville sediments are in general more quartz-rich than the Oronto Group sediments \cite{Hamblin1958a}.
